Quantifying the manageability of pollination networks in an invasion context

### 2017-05-04 paper/manuscript.Rmd

Disregarding practical considerations, any network (community) could hypothetically be fully controlled if we control the state of every single node (species) individually.
At the core of control theory of complex networks, however, rests the idea that the state of a node depends on the state of the nodes it interacts with, and the form of this dependence is determined by both the dynamic relationship among interacting nodes and the structure of the links in the network. 
This principle can therefore be harnessed to find a subset of *driver nodes* to which to apply external input signals which then drive the state of every other node in the network to a desired configuration [@Liu2016].

Conveniently the information necessary to determine the minimum number of driver nodes *D* in a system with linear dynamics is fully contained in the network structure [@Kalman1963; @Liu2011; @Motter2015]. 
Such a system can be described by $\frac{dx}{dt} = Ax + Bu(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example the species' abundances), an external time-varying input *u(t)* (the control vector), and two matrices *A* and *B*, which encapsulate information about the network structure and how the species respond to the external input, respectively. 
The matrix *B* has size $S \times D$, where *S* is the number of species in the community. 
Briefly, the focus of structural controllability, which we employ here, is to use the information contained in $\mathbf{A}$ to generate a supportable estimate of $\mathbf{B}$ (and by extension $D$).

This focus allows us to gain insight of the controllability of the network and the role of the species that compose it without being overly dependent on the particular choices of how the system dynamics are modelled or characterised. 
The trade-off is that, because of the assumption that ecological communities can be modeled with linear functional responses, at least near equilibrium points, structural controllability is not adequate to design a the time varying control signal *u(t)* that can drive the system from one particular equilibrium to another.	
Nevertheless, the lessons gained when assuming linearity—both at the network and the species level—are a prerequisite for understanding nonlinear control [@Liu2011; @Liu2016].

@@

In this study, we illustrate how a control-theoretic approach can be better adopted in network ecology to evaluate the effect of invasions and other kinds of perturbations. 
We study the different ways a management intervention can be structured and provide a starting point for the study controllability of ecological systems.
Although our pollination specific results might not be directly translatable to other ecological systems because of the degree constraints imposed by bipartite networks, the approach we propose is applicable wherever species abundances are influenced by their interactions, and exciting open questions lie ahead. 
How to design the precise "control signals" to reach a desired ecosystem state or conservation outcome? 
What are the implications of assuming fully nonlinear dynamics?
How important it is to include several interaction types for our understanding of manageability and species importance?
What are the implications for species coexistence? 
Which are the trade-offs between persistence at the species and the community level? 
Answering each of these questions might require its evaluation in different ecological systems, an explicit integration of control theory with numerical models of species densities [@Cornelius2013; @Gibson2016], and experimental tests on simple communities. 
Nevertheless, the potential rewards are encouraging from both an ecological and conservation perspective, where an integrated approach can shift the focus beyond the identification of ideal targets for intervention to the design of informed interventions that legitimately achieve restoration goals.

>>>>

Disregarding practical considerations, any network (community) could hypothetically be fully controlled if we control the state of every single node (species) individually.
At the core of control theory of complex networks, however, sits the idea that the state of any given node depends on the state of the nodes it interacts with, and the form of this dependence is determined by both the dynamic relationship among interacting nodes and the structure of the links in the network. 
This principle can be leveraged to find a subset of *driver nodes* to which to apply external input signals which then drive the state of every other node in the network to a desired configuration [@Liu2016].

Conveniently the information necessary to determine the minimum number of driver nodes *D* in a system with linear dynamics is fully contained in the network structure [@Kalman1963; @Liu2011; @Motter2015]. 
Such a system can be described by $\frac{dx}{dt} = Ax + Bu(t)$, where the change of its state over time ($\frac{dx}{dt}$) depends on its current state *x* (for example, the species' abundances), an external time-varying input *u(t)* (the control vector), and two matrices *A* and *B*, which encode information about the network structure and how the species respond to the external input, respectively. 
If *S* is the number of species in the community, the matrix *A* has size $S \times $S whereas the matrix *B* has size $S \times D$.

The goal of structural controllability, which we employ here, is to use the information contained in $\mathbf{A}$ to generate a supportable estimate of $\mathbf{B}$ (and by extension $D$).
This focus allows us to gain insight of the inherent controllability of a network, and the roles of the species that compose it, without being overly dependent on the particular choices of how the system dynamics are modelled or characterised. 
The trade-off is that, because of the assumption that ecological communities can be modeled with linear functional responses, structural controllability is not sufficient to fully design a the time-varying control signal *u(t)* that can drive the system from one particular equilibrium to another.	
Nevertheless, the lessons gained when assuming linearity---at both the network and the species level---are a prerequisite for understanding nonlinear control [@Liu2011; @Liu2016].

@@

In this study, we illustrate how a control-theoretic approach can be employed in network ecology to evaluate the effect of invasions and other kinds of perturbations. 
We study the different ways a management intervention can be structured and provide a starting point for the continued study of controllability in ecological contexts.
Although our pollination-specific results might not be directly translatable to other ecological networks that do not have bipartite structures, the approach we propose is applicable wherever species abundances are influenced by their interactions, and exciting open questions lie ahead. 
How to design the precise "control signals" to reach a desired ecosystem state or conservation outcome? 
What are the implications of assuming fully nonlinear dynamics?
How important it is to include several interaction types for our understanding of manageability and species importance?
What are the implications for species coexistence? 
Which are the trade-offs between persistence at the species and the community level? 
Answering each of these questions might require its evaluation in different ecological systems, an explicit integration of control theory with numerical models of species densities [@Cornelius2013; @Gibson2016], and experimental tests on simple communities. 
Nevertheless, the potential rewards are encouraging from both an ecological and conservation perspective, where an integrated approach can shift the focus beyond the identification of ideal targets for intervention to the design of informed interventions that legitimately achieve restoration goals.
